Fluent statistical computing interfaces for biological data analysis

### 2020-02-21 Rmd/03-fluentGenomics.Rmd
rebuild
update output

# Fluent genomics: integrating data with plyranges and tximeta {#ch:fluentGenomics}

@@

(ref:workflow) An overview of the fluent genomics workflow. First, we *import* data as a *SummarizedExperiment* object, which enables interoperability with downstream analysis packages. Then we *model* our assay data, using the existing Bioconductor packages *DESeq2* and *limma*. We take the results of our models for each assay with respect to their genomic coordinates, and *integrate* them. First, we compute the overlap between the results of each assay, then aggregate over the combined genomic regions, and finally summarize to compare enrichment for differentially expressed genes to non differentially expressed genes. The final output can be used for downstream visualization or further transformation.

(ref:maplot) Visualization of *DESeq2* results as an "MA plot". Genes that have an adjusted *p-value* below 0.01 are colored red.

(ref:linechart2) A line chart displaying how gene and peak counts change as the absolute DA LFC threshold increases. Lines are colored according to whether they represent a gene that is DE or not. Note the x-axis is on a $\log_{10}$ scale.

@@

Throughout the workflow (Figure \@ref(fig:workflow)), we will use existing
Bioconductor infrastructure to understand these datasets. In particular, we
will emphasize the use of the Bioconductor packages *plyranges* and *tximeta*.
The *plyranges* package fluently transforms data tied to genomic ranges using
operations like shifting, window construction, overlap detection, etc. It is
described by @Lee2019 and leverages underlying core Bioconductor infrastructure
[@granges; @bioc] and the *tidyverse* design principles @tidyverse.

The *tximeta* package described by @Love2019-tximeta is used to read RNA-seq
quantification data into R/Bioconductor, such that the transcript ranges and
their provenance are automatically attached to the object containing expression
values and differential expression results.

@@

The data used in this workflow is available from two packages: the *macrophage*
Bioconductor ExperimentData package and from the workflow package
*fluentGenomics*.

The *macrophage* package contains RNA-seq quantification from 24 RNA-seq
samples, a subset of the RNA-seq samples generated and analyzed by @alasoo. The
paired-end reads were quantified using *Salmon* [@salmon], using the Gencode 29
human reference transcripts [@gencode]. For more details on quantification, and
the exact code used, consult the vignette of the
[macrophage](http://bioconductor.org/packages/macrophage) package. The package
also contains the `Snakemake` file that was used to distribute the *Salmon*
quantification jobs on a cluster [@snakemake].

The *fluentGenomics* package contains functionality to download and generate a
cached *SummarizedExperiment* object from the normalized ATAC-seq data provided
by @alasooZenodo. This object contains all 145 ATAC-seq samples across all
experimental conditions as analyzed by @alasoo. The data can be also be
downloaded directly from the
[Zenodo](https://zenodo.org/record/1188300#.XIAhXlNKjOQ) deposition.

The following code loads the path to the cached data file, or if it is not
present, will create the cache and generate a *SummarizedExperiment* using the
the *BiocFileCache* package [@bcfilecache].

@@

### Using *tximeta* to import RNA-seq quantification data

@@

```{r dir, eval=FALSE}
dir <- "/path/to/quant/files"
```

where the path is relative to your current R session. However, in this case we
have distributed the files in the *macrophage* package. The relevant directory
and associated files can be located using `system.file`.

@@

Information about the experiment is contained in the `coldata.csv` file. We
leverage the *dplyr* and *readr* packages (as part of the *tidyverse*) to read
this file into R [@tidyverse]. We will see later that *plyranges* extends these
packages to accommodate genomic ranges.

@@

On a machine with a working internet connection, the above command works
without any extra steps, as the `tximeta` function obtains any necessary
metadata via FTP, unless it is already cached locally. The *tximeta* package
can also be used without an internet connection, in this case the linked
transcriptome can be created directly from a *Salmon* index and gtf.

@@

Because *tximeta* knows the correct reference transcriptome, we can ask
*tximeta* to summarize the transcript-level data to the gene level using the
methods of @Soneson2015.

@@

```{r coldata-atac, eval=FALSE}
atac_coldata <- read_tsv("ATAC_sample_metadata.txt.gz") %>%
  select(
    sample_id,
    donor,
    condition = condition_name
  ) %>%
  mutate(condition = relevel(factor(condition), "naive"))
```

@@

```{r mat-atac, eval=FALSE}
atac_mat <- read_tsv("ATAC_cqn_matrix.txt.gz",
                     skip = 1,
                     col_names =c("rownames", atac_coldata[["sample_id"]]))
rownames <- atac_mat[["rownames"]]
atac_mat <- as.matrix(atac_mat[,-1])
rownames(atac_mat) <- rownames
```

@@

```{r peaks-atac, eval=FALSE}
library(plyranges)
peaks_df <- read_tsv("ATAC_peak_metadata.txt.gz",
                     col_types = c("cidciicdc")
)

@@

```{r atac-se, eval=FALSE}
atac <- SummarizedExperiment(assays = list(cqndata=atac_mat),
                             rowRanges=peaks_gr,
                             colData=atac_coldata)
```

@@

We can easily run a differential expression analysis with *DESeq2* using the
following code chunks [@Love2014]. The design formula indicates that we want to
control for the donor baselines (`line`) and test for differences in gene
expression on the condition. For a more comprehensive discussion of DE
workflows in Bioconductor see @Love2016-f1000 and @Law2018-f1000.

@@

We now output the results as a *GRanges* object, and due to the conventions of
*plyranges*, we construct a new column called `gene_id` from the row names of
the results. Each row now contains the genomic region (`seqnames`, `start`,
`end`, `strand`) along with corresponding metadata columns (the `gene_id` and
the results of the test). Note that *tximeta* has correctly identified the
reference genome as "hg38", and this has also been added to the *GRanges* along
the results columns. This kind of book-keeping is vital once overlap operations
are performed to ensure that *plyranges* is not comparing across incompatible
genomes.

@@

For assessing differential accessibility, we run *limma* [@Smyth2004], and
generate the a summary of LFCs and adjusted p-values for the peaks:

@@

We now take the `rowRanges` of the *SummarizedExperiment* and attach the LFCs
and adjusted p-values from *limma*, so that we can consider the overlap with
differential expression. Note that we set the genome build to "hg38" and
restyle the chromosome information to use the "UCSC" style (e.g. "chr1",
"chr2", etc.). Again, we know the genome build from the Zenodo entry for the
ATAC-seq data.

@@

### Finding overlaps with *plyranges*

We have already used *plyranges* a number of times above, to `filter`,
`mutate`, and `select` on *GRanges* objects, as well as ensuring the correct
genome annotation and style has been used. The *plyranges* package provides a
grammar for performing transformations of genomic data [@Lee2019]. Computations
resulting from compositions of *plyranges* "verbs" are performed using
underlying, highly optimized range operations in the *GenomicRanges* package
[@granges].

@@

As *plyranges* is built on top of *dplyr*, it implements methods for many of
its verbs for *GRanges* objects. Here we can use `slice` to randomly sample the
rows of the `other_genes`. The `sample.int` function will generate random
samples of size equal to the number of DE-genes from the number of rows in
`other_genes`:

@@

Now we would like to modify our gene ranges so they contain the 10 kilobases on
either side of their transcription start site (TSS). There are many ways one
could do this, but we prefer an approach via the anchoring methods in
*plyranges*. Because there is a mutual dependence between the start, end,
width, and strand of a *GRanges* object, we define anchors to fix one of
`start` and `end`, while modifying the `width`. As an example, to extract just
the TSS, we can anchor by the 5' end of the range and modify the width of the
range to equal 1.

@@

We can then repeat the same pattern but this time using `anchor_center()` to
tell *plyranges* that we are making the TSS the midpoint of a range that has
total width of 20kb, or 10kb both upstream and downstream of the TSS.

@@

We are now ready to compute overlaps between RNA-seq genes (our DE set and
bootstrap sets) and the ATAC-seq peaks. In *plyranges*, overlaps are defined as
joins between two *GRanges* objects: a _left_ and a _right_ *GRanges* object.
In an overlap join, a match is any range on the _left_ *GRanges* that is
overlapped by the _right_ *GRanges*. One powerful aspect of the overlap joins
is that the result maintains all (metadata) columns from each of the _left_ and
_right_ ranges which makes downstream summaries easy to compute.

@@

@@

We have shown that by using *plyranges* and *tximeta* (with support of
Bioconductor and *tidyverse* ecosystems) we can fluently iterate through the
biological data science workflow: from import, through to modeling, and data
integration.

@@

The authors would like to thank all participants of the Bioconductor 2019 and
BiocAsia 2019 conferences who attended and provided feedback on early versions
of this workflow paper.

>>>>

# Fluent genomics with _plyranges_ and _tximeta_ {#ch:fluentGenomics}

@@

@@

(ref:workflow) An overview of the fluent genomics workflow. First, we *import* data as a *SummarizedExperiment* object, which enables interoperability with downstream analysis packages. Then we *model* our assay data, using the existing Bioconductor packages `DESeq2` and `limma`. We take the results of our models for each assay with respect to their genomic coordinates, and *integrate* them. First, we compute the overlap between the results of each assay, then aggregate over the combined genomic regions, and finally summarize to compare enrichment for differentially expressed genes to non differentially expressed genes. The final output can be used for downstream visualization or further transformation.

Throughout the workflow (Figure \@ref(fig:workflow)), we will use existing
Bioconductor infrastructure to understand these datasets. In particular, we
will emphasize the use of the Bioconductor packages `plyranges` and `tximeta`.
The `plyranges` package fluently transforms data tied to genomic ranges using
operations like shifting, window construction, overlap detection, etc. It is
described by @Lee2019 and leverages underlying core Bioconductor infrastructure
[@granges; @bioc] and the `tidyverse` design principles @tidyverse.

The `tximeta` package described by @Love2019-tximeta is used to read RNA-seq
quantification data into R/Bioconductor, such that the transcript ranges and
their provenance are automatically attached to the object containing expression
values and differential expression results.

@@

The data used in this workflow is available from two packages: the `macrophage`
Bioconductor ExperimentData package and from the workflow package
`fluentGenomics`.

The `macrophage` package contains RNA-seq quantification from 24 RNA-seq
samples, a subset of the RNA-seq samples generated and analyzed by @alasoo. The
paired-end reads were quantified using *Salmon* [@salmon], using the Gencode 29
human reference transcripts [@gencode]. For more details on quantification, and
the exact code used, consult the vignette of the
[macrophage](http://bioconductor.org/packages/macrophage) package. The package
also contains the `Snakemake` file that was used to distribute the *Salmon*
quantification jobs on a cluster [@snakemake].

The `fluentGenomics` package contains functionality to download and generate a
cached *SummarizedExperiment* object from the normalized ATAC-seq data provided
by @alasooZenodo. This object contains all 145 ATAC-seq samples across all
experimental conditions as analyzed by @alasoo. The data can be also be
downloaded directly from the
[Zenodo](https://zenodo.org/record/1188300#.XIAhXlNKjOQ) deposition.

The following code loads the path to the cached data file, or if it is not
present, will create the cache and generate a *SummarizedExperiment* using the
the `BiocFileCache` package [@bcfilecache].

@@

### Using `tximeta` to import RNA-seq quantification data

@@

```{r dir, eval=FALSE, echo = TRUE}
dir <- "/path/to/quant/files"
```

where the path is relative to your current R session. However, in this case we
have distributed the files in the `macrophage` package. The relevant directory
and associated files can be located using `system.file`.

@@

Information about the experiment is contained in the `coldata.csv` file. We
leverage the `dplyr` and `readr` packages (as part of the `tidyverse`) to read
this file into R [@tidyverse]. We will see later that `plyranges` extends these
packages to accommodate genomic ranges.

@@

On a machine with a working internet connection, the above command works
without any extra steps, as the `tximeta` function obtains any necessary
metadata via FTP, unless it is already cached locally. The `tximeta` package
can also be used without an internet connection, in this case the linked
transcriptome can be created directly from a *Salmon* index and gtf.

@@

Because `tximeta` knows the correct reference transcriptome, we can ask
`tximeta` to summarize the transcript-level data to the gene level using the
methods of @Soneson2015.

@@

```{r coldata-atac, eval=FALSE, echo = TRUE}
atac_coldata <- read_tsv("ATAC_sample_metadata.txt.gz") %>%
  select(
    sample_id,
    donor,
    condition = condition_name
  ) %>%
  mutate(condition = relevel(factor(condition), "naive"))
```

@@

```{r mat-atac, eval=FALSE, echo = TRUE}
atac_mat <- read_tsv("ATAC_cqn_matrix.txt.gz",
                     skip = 1,
                     col_names =c("rownames", atac_coldata[["sample_id"]]))
rownames <- atac_mat[["rownames"]]
atac_mat <- as.matrix(atac_mat[,-1])
rownames(atac_mat) <- rownames
```

@@

```{r peaks-atac, eval=FALSE, echo = TRUE}
library(plyranges)
peaks_df <- read_tsv("ATAC_peak_metadata.txt.gz",
                     col_types = c("cidciicdc")
)

@@

```{r atac-se, eval=FALSE, echo = TRUE}
atac <- SummarizedExperiment(assays = list(cqndata=atac_mat),
                             rowRanges=peaks_gr,
                             colData=atac_coldata)
```

@@

We can easily run a differential expression analysis with `DESeq2` using the
following code chunks [@Love2014]. The design formula indicates that we want to
control for the donor baselines (`line`) and test for differences in gene
expression on the condition. For a more comprehensive discussion of DE
workflows in Bioconductor see @Love2016-f1000 and @Law2018-f1000.

@@

(ref:maplot) Visualization of `DESeq2` results as an "MA plot". Genes that have an adjusted *p-value* below 0.01 are colored red.

We now output the results as a *GRanges* object, and due to the conventions of
`plyranges`, we construct a new column called `gene_id` from the row names of
the results. Each row now contains the genomic region (`seqnames`, `start`,
`end`, `strand`) along with corresponding metadata columns (the `gene_id` and
the results of the test). Note that `tximeta` has correctly identified the
reference genome as "hg38", and this has also been added to the *GRanges* along
the results columns. This kind of book-keeping is vital once overlap operations
are performed to ensure that `plyranges` is not comparing across incompatible
genomes.

@@

For assessing differential accessibility, we run `limma` [@Smyth2004], and
generate the a summary of LFCs and adjusted p-values for the peaks:

@@

We now take the `rowRanges` of the *SummarizedExperiment* and attach the LFCs
and adjusted p-values from `limma`, so that we can consider the overlap with
differential expression. Note that we set the genome build to "hg38" and
restyle the chromosome information to use the "UCSC" style (e.g. "chr1",
"chr2", etc.). Again, we know the genome build from the Zenodo entry for the
ATAC-seq data.

@@

### Finding overlaps with `plyranges`

We have already used `plyranges` a number of times above, to `filter`,
`mutate`, and `select` on *GRanges* objects, as well as ensuring the correct
genome annotation and style has been used. The `plyranges` package provides a
grammar for performing transformations of genomic data [@Lee2019]. Computations
resulting from compositions of `plyranges` "verbs" are performed using
underlying, highly optimized range operations in the `GenomicRanges` package
[@granges].

@@

As `plyranges` is built on top of *dplyr*, it implements methods for many of
its verbs for *GRanges* objects. Here we can use `slice` to randomly sample the
rows of the `other_genes`. The `sample.int` function will generate random
samples of size equal to the number of DE-genes from the number of rows in
`other_genes`:

@@

Now we would like to modify our gene ranges so they contain the 10 kilobases on
either side of their transcription start site (TSS). There are many ways one
could do this, but we prefer an approach via the anchoring methods in
`plyranges`. Because there is a mutual dependence between the start, end,
width, and strand of a *GRanges* object, we define anchors to fix one of
`start` and `end`, while modifying the `width`. As an example, to extract just
the TSS, we can anchor by the 5' end of the range and modify the width of the
range to equal 1.

@@

We can then repeat the same pattern but this time using `anchor_center()` to
tell `plyranges` that we are making the TSS the midpoint of a range that has
total width of 20kb, or 10kb both upstream and downstream of the TSS.

@@

We are now ready to compute overlaps between RNA-seq genes (our DE set and
bootstrap sets) and the ATAC-seq peaks. In `plyranges`, overlaps are defined as
joins between two *GRanges* objects: a _left_ and a _right_ *GRanges* object.
In an overlap join, a match is any range on the _left_ *GRanges* that is
overlapped by the _right_ *GRanges*. One powerful aspect of the overlap joins
is that the result maintains all (metadata) columns from each of the _left_ and
_right_ ranges which makes downstream summaries easy to compute.

@@

(ref:linechart2) A line chart displaying how gene and peak counts change as the absolute DA LFC threshold increases. Lines are colored according to whether they represent a gene that is DE or not. Note the x-axis is on a $log_{10}$ scale.

@@

We have shown that by using `plyranges` and `tximeta` (with support of
Bioconductor and `tidyverse` ecosystems) we can fluently iterate through the
biological data science workflow: from import, through to modeling, and data
integration.

@@

We would like to thank all participants of the Bioconductor 2019 and
BiocAsia 2019 conferences who attended and provided feedback on early versions of this workflow paper.

### 2020-02-21 index.Rmd
update output

```{r initial, echo = FALSE, cache = FALSE, results = 'hide'}
options("knitr.graphics.auto_pdf" = TRUE)
library(knitr)
opts_chunk$set(
  warning = FALSE, message = FALSE, echo = FALSE,
  fig.path = here::here('img/'), fig.align = 'center', fig.show = 'hold',
  cache = TRUE, external = TRUE, comment = "#>", dpi = 300,
  out.width = ifelse(is_html_output(), "100%", "\\textwidth"),
  cache.path = ifelse(is_html_output(), 
                      here::here('cache-gitbook/'), 
                      here::here('cache-pdfbook/')
  )
)
```

>>>>

```{r initial, echo = FALSE, cache = FALSE, results = 'hide'}
options("knitr.graphics.auto_pdf" = TRUE)
library(knitr)

opts_chunk$set(
  warning = FALSE, message = FALSE, echo = FALSE,
  fig.path = here::here('img/'), fig.align = 'center', fig.show = 'hold',
  cache = TRUE, external = TRUE, comment = "#>", dpi = 300,
  out.width = ifelse(is_html_output(), "100%", "\\textwidth"),
  cache.path = ifelse(is_html_output(), 
                      here::here('cache-gitbook/'), 
                      here::here('cache-pdfbook/')
  )
)

```
